Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-12-16 manuscript.Rmd
Changes

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Its examples include mixture models, latent class analysis with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Despite its popularity, there is a lack of standards for estimating and
reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
reporting guidelines for a specific type of LCA known as latent growth
models, general reporting guidelines for best practices in LCA are still
lacking.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address this gap in the literature by introducing updated guidelines for estimation and reporting on latent class analysis,
based on current best practice. 
Importantly, in order to lower the barrier of entry to latent class analysis
and ensure reproducibility of all examples,
this paper exclusively uses free, open source software for latent class analysis in R.
To this end, new functionality was developed in the R-package `tidySEM` [REF tidysem].

@@

Latent class analysis is a group of methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership. The concept of LCA can be understood in different ways.
A mixture model assumes that the study population is composed of $K$
subpopulations or classes.
It further assumes that the observed data are
a mixture of data originating from those subpopulations;
hence the name mixture model.
Consider the simplest possible
univariate "model", which is a normal distribution.
This model has
two parameters: the mean and the variance.
Latent class analysis aims to estimate the values of those parameters across a known number of classes $k$,
as well as the probability of belonging to classes $1 \ldots k$ for every individual $i$.
Commonly, the same model is estimated across all classes,
but with different parameters for each class (i.e., class-specific means and variances).
From this perspective, reporting the mean and variance of a variable is a special case of a mixture model with $k = 1$ class.

@@

Another way to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
A mixture model is like confirmatory factor analysis,
except that the continuous latent variable is substituted with a categorical latent variable.
One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes.
While factor analysis seeks to describe the relations between variables,
LCA seeks to describe unobserved groups and classify individual observations to those groups.
In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or individuals can be assigned to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach is probably acceptable.
This perspective on mixture modeling is sometimes referred to as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016]. 
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

@@

The term latent class analysis originally referred to cases where the
observed indicators were ordinal [@collins_latent_2009]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*.
Latent class models with ordinal indicators are parametrized differently from mixture models.
One common parametrization assumes that each categorical variable reflects an underlying standard normal distribution.
The parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.
Although this paper primarily focuses on latent class analysis with continuous indicators,
most of the suggested guidelines are equally applicable to ordinal indicators.

@@

Although latent class analysis is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that latent class analysis is similar to confirmatory factor analysis,
but with a categorical latent variable, it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
One challenge is that, unlike confirmatory factor analysis,
LCA does not provide absolute fit indices that can be used to perform inference.
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values.
A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.

@@

Extensive descriptive statistics (including the number of unique values,
variance of categorical variables, and missingness; see next subsection)
can be obtained using the function `tidySEM::descriptives(data)`. Note,
however, that sample-level descriptive statistics are of limited value
when the goal of a study is to identify subsamples using latent class
analysis. Plots (density plots for continuous variables, and bar charts
for categorical ones) may be more diagnostic. Note that density plots
can also aid in the choice of the number of classes, as further
explained in the section on visualization. Descriptive statistics and
plots can be relegated to online supplements, provided that these are
readily accessible [consider using a GitHub repository as a
comprehensive public research archive, as explained in
@van_lissa_worcs_2021].

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017].
While we concur that investigating missingness is due diligence,
it is important to emphasize that missingness is adequately handled by default in `OpenMx`, which is the computational backend of `tidySEM`.
Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
information without imputing missing values [@endersRelativePerformanceFull2001].
FIML is a best-practice solution for handling missing data;
on par with multiple imputation when the
estimation model includes the same variables as the imputation model
[@lee_comparison_2021].
Although FIML does not, by
default, handle missingness in exogenous variables - all indicator
variables in LCA are endogenous, so this is not a concern for LCA.
Multiple imputation, a common alternative to FIML,
is less suitable to LCA for two reasons. First,
because LCA methods are often computationally expensive,
and conducting them on multiple imputed datasets may be unfeasible.
Second, because many multiple imputation methods have some assumption of normality, which is likely to be violated when it is assumed that the data come from a heterogenous population.

Three types of missingness have
been distinguished in the literature [@rubin_inference_1976; 
@enders_applied_2022]: random missingness (MCAR);
missingness contingent on *observed* variables that can thus be known and controlled for (MAR);
and missingness related to unobserved variables, which can neither be known nor controlled for (MNAR).
A so-called "MCAR" test is often reported in relation to missingness
[@jamshidian_tests_2010].
But note that this name is somewhat misleading,
as a significant test indicates that missingness is MAR,
but a non-significant test statistic does not distinguish between MCAR or MNAR.
As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
This assumption is tenuous in the context of LCA.
A non-parametric MCAR test may be more suitable [@jamshidian_tests_2010].
For this tutorial,
the lead author has implemented this test in the `mice` package as `mice::mcar()`, with a
fast backend in C++ and new printing and plotting methods.
Our recommendation is to inspect missingness and report the proportion of missingness per variable using `tidySEM::descriptives()`,
and testing for MAR using `mice::mcar()`,
before proceeding with FIML.

One minor concern is that the K-means algorithm, which
`tidySEM` uses to determine starting values, is *not* robust to
partially missing data [@hartiganAlgorithm136KMeans1979].
When missing data causes it to fail,
`tidySEM` automatically switches to hierarchical clustering [@murtaghWardHierarchicalAgglomerative2014].
The user can also specify a different
clustering algorithm or specify starting values manually.
Cases with complete missing data should be removed,
or their data should be imputed another way.

@@

Before estimating a LCA model, researchers should carefully specify the model,
based on theoretical and empirical considerations.
This will ensure that the model accurately represents the data and provides reliable results.
Two factors play a role in LCA model specification:
the number of latent classes and the parametrization within each class.
The total number of parameters of a latent class model consists of the number of parameters within each class times the number of classes,
plus a classification vector with $K-1$ free elements.
Different types of latent class models have different parameters.
For example, latent profile analyses typically assume class-specific means,
equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`.
LCA with ordinal indicators typically assume free thresholds across classes,
where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`.
Like latent profile analyses, latent class growth analyses typically have free means across classes, but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`.
In the general sense, any structural equation model can be specified,
either using `lavaan` model syntax or using `OpenMx` RAM specification,
and supplied to the general function `mx_mixture()`.
The resulting model may have means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables.
All of these parameters can be freely estimated,
constrained to be equal across classes using equality constraints (e.g., `x ~~ a*x` fixes the variance of `x` to be equal across classes),
or fixed to a certain value (e.g., `x ~.5*1` fixes the mean of `x` to `.5`).

As any of these parameters could be misspecified,
prior literature has emphasized the importance of considering alternative model specifications [@van_de_schoot_grolts-checklist_2017].
While we agree with this recommendation,
two considerations have remained underemphasized in prior literature.
One consideration is that alternative model specifications may be approached differently depending on whether the LCA is data driven (exploratory), or theory driven
(confirmatory).
The section on Use Cases illustrates these different applications.

Prior literature has focused on exploratory LCA.
In this context, model specification typically consists of an exhaustive grid search along varying numbers of classes and model specifications.
The "correct" model specification is then determined based on a combination of fit indices, significance tests, and interpretability [@nylund_deciding_2007].
Confirmatory LCA does not require such an extensive search over the model space.
Theoretical models can be compared to a smaller set.
If hypothesized models have been specified a priori,
and possibly even preregistered,
it makes sense to focus on the estimation and interpretation of these models, rather than proceed in a purely data-driven manner.
One key challenge is that LCA lacks objective fit indices,
and the fit of a theoretical model can thus only be determined relative to other likely candidates.
For this reason, the theoretical model should still be compared to a few others.
A one-class model is a sensible benchmark, to test whether LCA is even appropriate.
A theoretical model would be supported by a substantial (and significant) difference in fit with a one-class model.
Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
Here again, a substantial difference in fit supports the theoretical model - but note that small differences in fit do not give cause to reject a theoretical model.
The theoretical model may still be preferred, for example, to ensure comparability with prior work.
The fact that the theoretical model received little differential support should of course be discussed,
along with potential implications for future work.
The BIC is the preferred fit index for comparing

@@

The second consideration is the risk of overfitting.
We address each in turn.

As explained before, the total number of parameters scales with the number of estimated classes.
Consequently, LCA methods have a potentially very high number of parameters.

@@

Many software packages are available for the estimation of LCA models.
Some of these packages have limited functionality, or implement specific
innovations. Other packages implement LCA in the context of a more
flexible structural equation modeling framework. The most notable
examples of the latter are the commercial programs Mplus and Latent
GOLD, and the free open source R-package OpenMx. The commercial packages
stand out because they offer relatively user-friendly interfaces and
implement sensible defaults for complex analyses, including LCA methods.
This lowers the threshold for applied researchers to adopt such methods.
Commercial software also has several downsides, however. One such
downside is that use of the software is restricted to those individuals
and institutions who can afford a license. Another downside is that the
source code, being proprietary, cannot be audited, debugged, or enhanced
by third parties. This incurs the risk that mistakes in the source code
may go unnoticed, and curbs progress as software developers cannot add
new functionality.

@@

LCA parameters and model fit statistics can be estimated in a variety of
ways. The choice of the estimator depends on the presence of missing
values, sample size, number of indicators, and available computational
resources [@weller_latent_2020]. A commonly used technique is maximum
likelihood (ML) estimation with the expectation-maximization (EM)
algorithm as a local optimizer. Imagine we are estimating two
parameters, e.g. the class-specific means $\mu_c$ on a continuous
indicator (ignoring the variance for now). The EM algorithm will attempt
to find a combination of values for these two parameters that maximizes
the likelihood ($LL$) of all observed data. In practice, instead of
maximizing $LL$, often $-2*LL$ is minimized, as this offers
computational advantages. We can think of this optimization problem as a
three-dimensional landscape: The X and Y dimensions are determined by
the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the
Z-dimension is determined by $Z = -2*LL$. The optimizer must find the
deepest "valley" in this landscape, which reflects the combination of
$\mu_1$ and $\mu_2$ that maximizes the likelihood of the data. The EM
optimizer behaves somewhat like a marble dropped in this landscape. It
is dropped at some random point in space, and will roll into the nearest
valley. The problem is that, once EM rolls into a valley, it will settle
on the bottom of that valley (this is known as "convergence"). It cannot
climb out again. Thus, if their are multiple valleys, the risk is that
the optimizer gets stuck in a shallower valley (a "local optimum"), and
never discovers the deepest valley (the "global optimum", or best
solution). One solution to this problem is to drop many marbles at
random places, compare their final $-2*LL$ values, choose the solution
with the lowest $-2*LL$, and make sure that several marbles replicated
this solution. This is the "random starts" approach.

>>>>

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Examples include mixture models, latent class analysis with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Despite its popularity, there is a lack of standards for estimating and
reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
reporting guidelines for a specific type of LCA known as latent growth
models, general reporting guidelines for best practices in LCA are still
lacking.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address this gap in the literature by introducing updated guidelines for estimation and reporting on latent class analysis,
based on current best practice. 
Importantly, in order to lower the barrier of entry to latent class analysis
and ensure reproducibility of all examples,
this paper exclusively uses free, open source software for latent class analysis in R.
To this end, new functionality was developed in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022].

@@

Latent class analysis is a group of methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership [@vermuntLatentClassAnalysis2004].
This method has become quite popular across scientific fields,
under a number of different names;
most notably (finite Gaussian) mixture modeling and latent profile analysis.
The concept of LCA can be understood in different ways.
Generally speaking, LCA assumes that the study population is composed of $K$
subpopulations.
It further assumes that the observed data are
a mixture of data originating from those subpopulations;
hence the name mixture model.
Consider the simplest possible
univariate "model", which is a normal distribution.
This model has
two parameters: the mean and the variance.
LCA aims to estimate the values of those parameters across a known number of classes $K$,
as well as the probability of every individual $i$ belonging to classes $1 \ldots k$.
Commonly, the same model is estimated across all classes,
but with different parameters for each class (i.e., class-specific means and variances).

@@

Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
A mixture model is like confirmatory factor analysis,
except with a categorical instead of a continuous latent variable.
One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes.
While factor analysis seeks to describe the relations between variables,
LCA seeks to describe unobserved groups and classify individual observations to those groups.
In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or individuals can be assigned to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
This perspective on mixture modeling is sometimes referred to as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016]. 
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

@@

The term latent class analysis originally referred to cases where the
observed indicators were ordinal (ordered categorical) [@collins_latent_2009]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*.

<!-- Although this paper primarily focuses on latent class analysis with continuous indicators, -->
<!-- most of the suggested guidelines are equally applicable to ordinal indicators. -->

@@

Although latent class analysis is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that latent class analysis is similar to confirmatory factor analysis,
but with a categorical latent variable, it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellanResilientOvercontrolledUndercontrolled2010].
One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices.
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values.
A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.

@@

Extensive descriptive statistics (including the number of unique values,
variance of categorical variables, and missingness; see next subsection)
can be obtained using `tidySEM::descriptives(data)`. Note,
however, that sample-level descriptive statistics are of limited value
when the goal of a study is to identify subsamples using latent class
analysis. Plots (density plots for continuous variables, and bar charts
for categorical ones) may be more diagnostic. Note that density plots
can also aid in the choice of the number of classes, as further
explained in the section on visualization. Descriptive statistics and
plots can be relegated to online supplements, provided that these are
readily accessible [consider using a GitHub repository as a
comprehensive public research archive, as explained in
@van_lissa_worcs_2021].

### Data preprocessing

LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate means and (co)variances.
Note that this assumption is likely violated for Likert-type scales.
For the remaining three measurement levels,
some data preprocessing is required.
The exact data types accepted by `tidySEM` at the time of writing are `numeric` and `ordered`.
To correctly specify ordered variables,
such as Likert scales
use the function `OpenMx::mxFactor()`.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal.
Finally, nominal variables are converted to binary indicators of group membership via dummy coding, for example, using `mx_dummies()`.

A second consideration in data preprocessing is that models with indicators on different scales may present with convergence issues.
The reason for this is that the parameter space is high-dimensional with many potential local optima.
If one indicator is on a much larger scale than another, the parameter space may be more extensively explored for that parameter.
The problem can be resolved by transforming the indicators to place them on (roughly) similar scales.
The most common transformation is standardization, which involves subtracting the mean of each variable from each value and dividing by the standard deviation.
In R, this transformation is applied using the `scale()` function.
Standardization retains all available information,
and model parameters can be transformed back to the original scale of the indicators by reversing the transformation.
The transformed variable has a mean of zero and a standard deviation of one.
Note that standardization is typically not necessary if indicators are already on the same, or similar, scales.

@@

Previous work has emphasized the importance of examining the pattern of
missing data and reporting how missingness was handled
[@van_de_schoot_grolts-checklist_2017].
While we concur that investigating missingness is due diligence,
it is important to emphasize that missingness is adequately handled by default in `OpenMx`, which is the computational backend of `tidySEM`.
Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
information without imputing missing values [@endersRelativePerformanceFull2001].
FIML is a best-practice solution for handling missing data;
at least on par with multiple imputation [@lee_comparison_2021].
Multiple imputation is less suitable to LCA for two reasons. First,
because LCA methods are often computationally expensive,
and conducting them on multiple imputed datasets may be unfeasible.
Second, because many multiple imputation methods have some assumption of normality, which is likely to be violated when it is assumed that the data come from a heterogenous population.
FIML is not suitable for cases with complete missing data.

Three types of missingness have
been distinguished in the literature [@rubin_inference_1976; 
@enders_applied_2022]: random missingness (MCAR);
missingness contingent on *observed* variables that can thus be known and controlled for (MAR);
and missingness related to unobserved variables, which can neither be known nor controlled for (MNAR).
A so-called "MCAR" test is often reported in relation to missingness
[@jamshidian_tests_2010].
But note that this name is somewhat misleading,
as a significant test indicates that missingness is MAR,
but a non-significant test statistic does not distinguish between MCAR or MNAR.
As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
This assumption is tenuous in the context of LCA.
A non-parametric MCAR test may be more suitable [@jamshidian_tests_2010].
For this tutorial,
the lead author has implemented this test in the `mice` package as `mice::mcar()`.

To conclude; our recommendation is to inspect and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
and to test for MAR using `mice::mcar()`.
As FIML estimation assumes that missingness is either MCAR or MAR,
one would proceed with FIML regardless of the outcome of the test.
One minor concern is that the approach `tidySEM` uses to determine starting values is *not* robust to
partially missing data.
This should not be a problem, however, as uses a global optimizer to estimate LCA models,
which should be independent of starting values [@coranaMinimizingMultimodalFunctions1987].

@@

Any LCA model should be carefully specified based on theoretical and empirical considerations.
The parameters of LCA depend on the measurement level of the indicators.
LCA with continuous indicators can be parametrized in terms of means, variances, and covariances.
Latent class growth analyses have the same parameters,
but with respect to the latent growth variables.
LCA with ordinal indicators can be parametrized differently.
The approach implemented in `tidySEM` assumes that each ordinal variable reflects an underlying standard normal distribution.
The parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$).
These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.
All of the aforementioned parameters can be freely estimated,
or constrained across classes, or fixed (e.g., to zero).

Prior literature emphasized the importance of considering alternative model specifications [@vandeschootGRoLTSChecklistGuidelinesReporting2017].
However, what has remained underemphasized is that alternative model specifications may be approached differently depending on whether an analysis is exploratory or confirmatory (see Use Cases).
Most prior literature has emphasized exploratory applications of LCA [e.g., @nylund_deciding_2007]
In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
A "final" model is then selected based on a combination of fit indices,
significance tests,
and interpretability (see Class Enumeration).

Confirmatory LCA does not require such an extensive search over the model space.
If hypothesized models have been specified a priori,
and possibly even preregistered,
it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
Nonetheless, an important reason to consider alternative model specifications is that LCA lacks objective fit indices.
The fit of a theoretical model can thus only be assessed by comparison to other candidates.
For this reason, the theoretical model should still be compared to a few others.
A one-class model is a sensible benchmark, to test whether LCA is even appropriate.
Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
If the theoretical model has much better fit than other models,
this provides evidence for the theory.
If differences in fit are small, however, this may not  do not give cause to reject a theoretical model.
The theoretical model may still be preferred, for example, to ensure comparability with prior work.
The fact that the theoretical model received little differential support should of course be discussed,
along with potential implications for future work.

<!-- This will ensure that the model accurately represents the data and provides reliable results. -->
<!-- Two factors play a role in LCA model specification: -->
<!-- the number of latent classes and the model estimated within each class. -->
<!-- Assuming the same model is estimated across classes, -->
<!-- the total number of parameters in LCA is equal to the number of parameters of the class-specific model times the number of classes, -->
<!-- plus a vector of class proportions with $K-1$ free elements. -->

<!-- In the general sense, the class-specific model can be any structural equation model. -->
<!-- Its parameters may include means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables. -->
<!-- All of these parameters can be freely estimated, -->
<!-- constrained to be equal across classes, -->
<!-- or fixed to a known value. -->
<!-- The specific types LCA introduced in the Taxonomy all have different parameters. -->
<!-- For example, LPA typically assumes class-specific means, -->
<!-- equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`. -->
<!-- By contrast, LCA with ordinal indicators assumes free thresholds across classes, -->
<!-- where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`. -->
<!-- Like LPA, LCGA and GMM typically have free means across classes, -->
<!-- but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`. -->
<!-- The distinction between LCGA and GMM is whether variances of the latent growth variables are constrained to zero, or freely estimated (either across or within classes). -->
<!-- either using `lavaan` model syntax or using `OpenMx` RAM specification, -->
<!-- and supplied to the general function `mx_mixture()`. -->
<!-- The resulting model may have -->

@@

A second consideration in model specification is the risk of overfitting [@hastieElementsStatisticalLearning2009].
Overfitting means that a model captures idiosyncratic noise in a sample,
and thus generalizes poorly to new samples.
It occurs when a model has many parameters relative to the number of observations
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number of parameters.

In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseauAsymptoticBehaviourPosterior2011].
Although it is not known whether frequentist LCA exhibits the same behavior,
this suggests that low class counts might indicate overfitting.

@@

LCA models can be estimated in a variety of ways.
Some of these packages have limited functionality, or implement specific
innovations. Other packages implement LCA in the context of a more
flexible structural equation modeling framework. The most notable
examples of the latter are the commercial programs Mplus and Latent
GOLD, and the free open source R-package OpenMx. The commercial packages
stand out because they offer relatively user-friendly interfaces and
implement sensible defaults for complex analyses, including LCA methods.
This lowers the threshold for applied researchers to adopt such methods.
Commercial software also has several downsides, however. One such
downside is that use of the software is restricted to those individuals
and institutions who can afford a license. Another downside is that the
source code, being proprietary, cannot be audited, debugged, or enhanced
by third parties. This incurs the risk that mistakes in the source code
may go unnoticed, and curbs progress as software developers cannot add
new functionality.

@@

LCA parameters and model fit statistics can be estimated in a variety of
ways. The choice of the estimator depends on the presence of missing
values, sample size, number of indicators, and available computational
resources [@weller_latent_2020]. A commonly used technique is maximum
likelihood (ML) estimation with the expectation-maximization (EM)
algorithm as a local optimizer. Imagine we are estimating two
parameters, e.g. the class-specific means $\mu_c$ on a continuous
(ignoring the variance for now).
EM attempts to find a combination of values for these two parameters that maximizes the log likelihood ($LL$) of the data.
In practice, instead of maximizing $LL$,
$-2*LL$ is minimized.
Transforming the likelihood this way does not change the resulting parameter estimates,
but it is mathematically convenient.
Moreover, the $-2*LL$ is asymptotically $\chi^2$ distributed if the null-hypothesis is true, and thus this quantity can be used for model fit tests [@wilksLargeSampleDistributionLikelihood1938].
To understand the optimization problem, imagine a three-dimensional landscape: 
The X and Y dimensions represent potential values of the class-specific means, $X = \mu_1$ and $Y = \mu_2$, and the Z-dimension is determined by $Z = -2*LL$. 
The optimizer must find the
deepest "valley" in this landscape,
which reflects the combination of $\mu_1$ and $\mu_2$ that minimizes the $-2*LL$.
The EM
optimizer behaves somewhat like a marble dropped in this landscape. It
is dropped at some random point in space, and will roll into the nearest
valley.
The challenge is that LCA models are often highly parametrized.
This means that the landscape is not three-dimensional, but (very) high-dimensional,
with potentially many valleys and sparse data to provide information about which valley is deepest.
If the marble rolls into one of these valleys,
it will settle there and not climb out again.
The estimator has "converged".
The risk is that the optimizer gets stuck in a shallower valley (a "local optimum"),
and never discovers the deepest valley (the "global optimum", or best solution).
One solution to this problem is to drop many marbles at random places,
compare their final $-2*LL$ values,
choose the solution
with the lowest $-2*LL$, and make sure that several marbles replicated
this solution.
This is the "random starts" approach implemented in, e.g., Mplus.

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values,
it is inefficient because many marbles are likely to start very far away from any valley.
Finally, if all sets of starting values are close to a local optimum,
there is a risk that the global optimum is not discovered.
An alternative solution that overcomes these challenges is to use a global optimizer, like the simulated annealing algorithm [SA, @coranaMinimizingMultimodalFunctions1987].
SA does not behave quite like the marbles of the random starts approach;
instead, it iteratively considers some "destination" in the landscape,
and compares its $-2*LL$ to the current one.
If the destination $-2*LL$ is lower, 
the estimator moves there.
However, if the destination  $-2*LL$ is *higher*,
the estimator still moves there occasionally, with a certain probability.
Think of this as occasionally flicking the marble back out of valleys,
to see if it rolls back into the current valley, or finds another,
even deeper one.
This property allows SA it to escape local optima and find the global optimum.
